# Wiring Up Drill and R (`dplyr`-style)
```{r message=FALSE, include=FALSE, warning=FALSE}
library(sergeant)
library(tidyverse)
```

## Problem

You have Drill installed and want to work with Drill from R using `dplyr` idioms.

## Solution

Install, load and use the `sergeant` package.

## Discussion

Working with Drill in a `dplyr` context is pretty straightforward. Assuming you have drill running you need to:

- connect to the database
- identify a table
- perform normal operations

The Drill query interface (<http://localhost:8047/query>) provides an example query we can use for testing if your setup is working:

    SELECT * FROM cp.`employee.json` LIMIT 20

Let's take a look at the `employee.json` table using R & `dplyr`.

```{r 02-example-01-connect, message=FALSE, warning=FALSE, cache=TRUE}
library(sergeant)
library(tidyverse)

db <- src_drill("localhost")

db

employee <- tbl(db, "cp.`employee.json`")

employee
```

If you've never worked with the `dplyr` and databases before, you may not be aware that what's really happening is different than what you experince with local R data frames.

The call to `src_drill("localhost")` does some RJDBC/DBI "magic" behind the scenes to setup all the necessary connection parameters.

By printing `db`, a SQL query ---`SHOW DATABASES`---is issued to get available storage/databases and reformats it to conform to the expected `dplyr` datbase API.

When `tbl(...)` is executed a

    SELECT * FROM  cp.`employee.json`  LIMIT 1

is issued to obtain field information for future operations on `employee`.

Finally, when `employee` is printed, the query

    SELECT * FROM  cp.`employee.json`  LIMIT 10

is issued and the contents transferred back into R. 

Notice that the data is really all still in Drill. Future recipes introduce `dplyr::collect()` which is how you ultimately transfer the query results back into an R object.

## See Also

- [Drill package source README](https://github.com/hrbrmstr/sergeant/blob/master/README.md)